---
title: "Calculating delta Zscores for immune cell modules and making plots"
---
# Adaptation of original cytokine modules script by Dr Venturini to immune cell modules

In this script, I will firstly be calculating the mean Z-score for each immune cell module across the 12 normal samples. I will then calculate the delta Z-scores (difference from mean normal Z-score) for each normal, brain, AAV2, gene-therapy and seronegative hepatitis sample. Finally I will create plots to visualise differences between groups.

```{r}
library(tidyverse)
library(rstatix)
library(ggpubr)
library(stringr)
```


=========================== PART 1: calculate mean Z-scores across normal samples =================================

Load the file with normal liver data
```{r}
GSE142530_healthyLiver <- read.csv("path/to/immune/cell/module/Zscores/for/normal/liver.csv")
dim(GSE142530_healthyLiver)
```

Change it into the correct format for calculations
```{r}
GSE142530_healthyLiver_long <- GSE142530_healthyLiver %>% 
  pivot_longer(3:14, names_to = "Sample", values_to = "Zscore") %>% 
  mutate("Class"="Normal") %>%
  rename("Module"="cell_type") %>% 
   select(-"X")
```

Calculate mean z-scores for normal
```{r}
means_normal <- GSE142530_healthyLiver_long %>%  
  group_by(Module,Class) %>% 
  summarise(mean(Zscore)) %>% 
  filter(Class=="Normal") %>% 
  select(-Class)
```

===================================== PART 2: load data for other samples ====================================

Load AAV2 liver Z-scores and reformat data
```{r}
zscores_AAV2 <- read_csv("path/to/immune/cell/module/Zscores/for/AAV2/liver.csv")

zscores_AAV2_long <- zscores_AAV2 %>% 
  pivot_longer(3:6, names_to = "Sample", values_to = "Zscore") %>% 
  mutate("Class"="AAV2") %>%
  rename("Module"="cell_type") %>% 
   select(-"...1")
```

Load brain control Z-scores and reformat data
```{r}
zscores_brain <- read_csv("path/to/immune/cell/module/Zscores/for/brain/control.csv")

brain <- zscores_brain %>% 
  mutate("Class"="Brain") %>% #class name can change 
  rename("Module"=cell_type, "Zscore"=xxxxx.xxxxx104) %>% #CHANGE HERE!
  select(-"...1") %>% # removes the ...1 column
  mutate("Sample"="xxxxx.xxxxx104") #CHANGE HERE!
```

Load xxxxx.xxxxx159 gene therapy sample Z-scores and reformat data
```{r}
zscores_gene_therapy <- read_csv("path/to/immune/cell/module/Zscores/for/gene/therapy/sample.csv")

gene_therapy <- zscores_gene_therapy %>% 
  mutate("Class"="Gene Therapy") %>% #class name can change 
  rename("Module"=cell_type, "Zscore"=xxxxx.xxxxx159) %>% #CHANGE HERE!
  select(-"...1") %>% # removes the ...1 column
  mutate("Sample"="xxxxx.xxxxx159") #CHANGE HERE!
```

Load xxxxx.xxxxx035 Z-scores and reformat data 
```{r}
zscores_xxxxxxxxxx035 <- read_csv("path/to/immune/cell/module/Zscores/for/sample/xxxxxxxxxx035.csv")
dim(zscores_xxxxxxxxxx035)

zscores_xxxxxxxxxx035_long <- zscores_xxxxxxxxxx035 %>% 
  pivot_longer(3, names_to = "Sample", values_to = "Zscore") %>% 
  mutate("Class"="xxxxx.xxxxx035") %>%
  rename("Module"="cell_type") %>% 
   select(-"...1")
```

Load xxxxx.xxxxx045 Z-scores and reformat data 
```{r}
zscores_xxxxxxxxxx045 <- read_csv("path/to/immune/cell/module/Zscores/for/sample/xxxxxxxxxx045.csv")
dim(zscores_xxxxxxxxxx045)

zscores_xxxxxxxxxx045_long <- zscores_xxxxxxxxxx045 %>% 
  pivot_longer(3, names_to = "Sample", values_to = "Zscore") %>% 
  mutate("Class"="xxxxx.xxxxx045") %>%
  rename("Module"="cell_type") %>% 
   select(-"...1")
```



========================== PART 3: putting data together and calculating delta Z-scores =====================

Join data together and adjust z-scores with the "normal" group.
```{r}
e <- rbind(GSE142530_healthyLiver_long, zscores_AAV2_long, brain, gene_therapy, zscores_xxxxxxxxxx035_long, zscores_xxxxxxxxxx045_long)
f <- e %>% 
  left_join(means_normal, by="Module") %>% 
  rename("NormalLiverAve.Zscore"="mean(Zscore)") %>% 
  mutate("Delta.Zscore"=Zscore-NormalLiverAve.Zscore)
```

```{r}
subset_modules_ontogeny <- c("T_cell", "NK", "myeloid")
ontogeny <- f %>% filter(Module %in% subset_modules_ontogeny)
```

```{r}
subset_modules_Tcell_subsets <- c("CD4_T", "CD8_T", "atypical_CD8", "gd_T", "NKT")
Tcell_subsets <- f %>% filter(Module %in% subset_modules_Tcell_subsets)
```

```{r}
subset_modules_Tcell_functions <- c("T1", "T22", "Treg", "naive", "cytotoxic")
Tcell_functions <- f %>% filter(Module %in% subset_modules_Tcell_functions)
```

```{r}
subset_modules_myeloid_functions <- c("antimicrobial", "Ag_presenting")
myeloid_functions <- f %>% filter(Module %in% subset_modules_myeloid_functions)
```

Rename modules
```{r}
ontogeny$Module[ontogeny$Module == "T_cell"] <- "T cell"
Tcell_subsets$Module[Tcell_subsets$Module == "CD4_T"] <- "CD4"
Tcell_subsets$Module[Tcell_subsets$Module == "CD8_T"] <- "CD8"
Tcell_subsets$Module[Tcell_subsets$Module == "atypical_CD8"] <- "Atypical CD8"
Tcell_subsets$Module[Tcell_subsets$Module == "gd_T"] <- "Gamma delta"
Tcell_subsets$Module[Tcell_subsets$Module == "NKT"] <- "NKT"
ontogeny$Module[ontogeny$Module == "myeloid"] <- "Myeloid cell"
Tcell_functions$Module[Tcell_functions$Module == "Treg"] <- "Regulatory"
Tcell_functions$Module[Tcell_functions$Module == "cytotoxic"] <- "Cytotoxic"
Tcell_functions$Module[Tcell_functions$Module == "naive"] <- "Naive"
myeloid_functions$Module[myeloid_functions$Module == "antimicrobial"] <- "Antimicrobial"
myeloid_functions$Module[myeloid_functions$Module == "Ag_presenting"] <- "Ag-presenting"
ontogeny$Module[ontogeny$Module == "NK"] <- "NK cell"
```


========================================= PART 4: PLOTS ==============================================

```{r}
library(RColorBrewer) 
display.brewer.all()
display.brewer.pal(n= 9, name= "Set1")  
brewer.pal(n = 9, name = "Set1")  
```

Create My_Theme
```{r}
t = 10 #size of text
m = 8 #size of margin around text
tc = "black" #colour of text
My_Theme = theme(
  axis.title.x = element_text(size = 13, face = "bold", margin = margin(t = m)),
  axis.text.x = element_text(size = 13, face = "bold", colour = tc, angle = 0, hjust = 0.5),
  axis.title.y.left = element_text(size = 13, face = "bold", margin = margin(r = m)),
  axis.title.y.right = element_text(size = 13, face = "bold", margin = margin(l = m)),
  axis.text.y = element_text(size = 13, face = "bold", colour = tc),
  legend.title = element_text(size=13, face = "bold", colour = tc),
  legend.text = element_text(size=13, face = "bold", colour = tc),
  plot.title = element_text(size=t, face = "bold", colour = tc),
  strip.text = element_text(size=t, face = "bold", colour = tc),
  strip.background = element_rect(fill = "gray90", colour = "black", size = 0.5),
  panel.border = element_rect(fill = NA, linewidth = 0.5, colour = tc),
  panel.background = element_rect(fill = "gray97"),
  plot.background = element_rect(fill='transparent', color=NA), #transparent plot bg
  legend.position = "right", legend.justification = "top"
)

ontogeny$Class <- factor(ontogeny$Class, levels = c("Normal", "AAV2", "Gene Therapy", "xxxxx.xxxxx035", "xxxxx.xxxxx045", "Brain"))
Tcell_subsets$Class <- factor(Tcell_subsets$Class, levels = c("Normal", "AAV2", "Gene Therapy", "xxxxx.xxxxx035", "xxxxx.xxxxx045", "Brain"))
Tcell_functions$Class <- factor(Tcell_functions$Class, levels = c("Normal", "AAV2", "Gene Therapy", "xxxxx.xxxxx035", "xxxxx.xxxxx045", "Brain"))
myeloid_functions$Class <- factor(myeloid_functions$Class, levels = c("Normal", "AAV2", "Gene Therapy", "xxxxx.xxxxx035", "xxxxx.xxxxx045", "Brain"))
```


### Immune cell ontogeny

Make scatterplot
```{r}
ggplot(ontogeny)+
  geom_point(aes(Module, Delta.Zscore, colour = Class), size = 3, alpha = 0.8)+
  scale_colour_manual(values = c("#92a0b3", "#de102a", "#ff7821", "#F781BF", "#b87dc1", "#4DAF4A"))+
  geom_hline(aes(yintercept = 0), linewidth = 0.5, linetype = "dashed")+
  scale_y_continuous(limits=c(-2.5, 6), breaks=c(-2.5, 0, 2.5, 5))+
  labs(colour = "Group",
       y = "Delta Zscore cf. average of normal liver", 
       x = "Immune cell ontogeny modules") +
  My_Theme + theme(axis.text.x = element_text(angle = -45, hjust=0))
```


Make PCA plot
```{r}
library(broom)

### calculating PCA data
mydata_forpca1 <- ontogeny %>% select(Module, Class, Sample, Delta.Zscore) %>% 
  pivot_wider(names_from = Module, values_from = Delta.Zscore)

pca_calc1 <- mydata_forpca1 %>% select(where(is.numeric)) %>% prcomp(scale = F)

### plotting PCA
pca_calc1 %>% augment(mydata_forpca1) %>%  ggplot(aes(.fittedPC1, .fittedPC2, color = Class))+ 
  geom_point(size = 2)+
  scale_color_manual(values=c(Normal="#92a0b3",AAV2="#de102a", "Gene Therapy"="#ff7821", "xxxxx.xxxxx035"="#F781BF", "xxxxx.xxxxx045"="#b87dc1", Brain= "#4DAF4A"))+
  labs(x="PC1",y="PC2")+
  My_Theme
```

Make heatmap
```{r}
library(ComplexHeatmap)
library(circlize)

### create numerical data matrix with row & column names
t=12
heatmapData1 <- mydata_forpca1 %>% select(-Class) %>%
  remove_rownames %>% column_to_rownames(var="Sample") %>% t

### create annotations (for help see: https://www.biostars.org/p/317349/)
FilenameAnnotations <- ontogeny %>% select(Class,Sample) %>% distinct()
i <- FilenameAnnotations
ann <- data.frame(i$Class)
colnames(ann) <- c("Class")
ann$Class <- factor(ann$Class, levels = c("Normal","AAV2", "Gene Therapy", "xxxxx.xxxxx035", "xxxxx.xxxxx045", "Brain"))
colours <- list("Class" = c(
  "Normal"=	"#92a0b3",
  "AAV2"=	"#de102a",
  "Gene Therapy"=	"#ff7821",
  "xxxxx.xxxxx035"="#F781BF",
  "xxxxx.xxxxx045"="#b87dc1",
  "Brain" = "#4DAF4A"))
colAnn <- HeatmapAnnotation(df = ann,
                            which = 'col',
                            col = colours,
                            border = T,
                            show_annotation_name=F,
                            annotation_name_gp = gpar(fontsize=8, fontface = "bold")
                            )

### select heatmap colours
library(colorRamp2)
col_fun = colorRamp2(c(0, 5, 10), c("#2166AC", "#FDF0D5", "#950018"))

### plot heatmap 
hm1 <- Heatmap(heatmapData1,
              col = col_fun,
              border_gp = gpar(col = "black", lty = 1),
              rect_gp = gpar(col = "black", lwd = 1),
              #column_title = "...",
              #column_title_gp = gpar(fontsize = 10),
              row_title = "Immune cell ontogeny modules",
              row_title_gp = gpar(fontsize=12, fontface = "bold"),
              show_row_names = T,
              row_names_gp = gpar(fontsize=10, fontface = "bold"),
              show_column_names = F,
              column_title = "Samples",
              column_title_gp = gpar(fontsize=12, fontface = "bold"),
              #column_names_gp = gpar(fontsize=t),
              #column_names_rot = 0,
              #column_names_centered = T,
              #column_names_gp = gpar(fontsize = 10),
              #name = "z score",
              cluster_rows = T,
              cluster_columns = T,
              clustering_method_columns = "ward.D",
              column_split = 2,
              #row_split = 2,
              top_annotation = colAnn,
              heatmap_height = unit(7, "cm"),
              heatmap_width = unit(14, "cm"),
              #right_annotation = row.ha,
              #show_annotation_name = F,
              heatmap_legend_param = list(
                direction = "horizontal",
                #grid_height = 0.5,
                at = c(0, 2, 4, 6, 10),
                labels = c(-0, 2, 4, 6, 10),
                title = "Expression delta Z-score",
                legend_width = unit(2, "cm"),
                title_position = "lefttop")
              )

draw(hm1, heatmap_legend_side = "bottom", annotation_legend_side = "right")
```


### T cell subsets

Make scatterplot
```{r}
ggplot(Tcell_subsets)+
  geom_point(aes(Module, Delta.Zscore, colour = Class), size = 3, alpha = 0.8)+
  scale_colour_manual(values = c("#92a0b3", "#de102a", "#ff7821", "#F781BF", "#b87dc1", "#4DAF4A"))+
  geom_hline(aes(yintercept = 0), linewidth = 0.5, linetype = "dashed")+
  scale_y_continuous(limits=c(-2.5, 6), breaks=c(-2.5, 0, 2.5, 5))+
  labs(colour = "Group",
       y = "Delta Zscore cf. average of normal liver", 
       x = "T cell subset modules") +
  My_Theme + theme(axis.text.x = element_text(angle = -45, hjust=0))
```


Make PCA plot
```{r}
library(broom)

### calculating PCA data
mydata_forpca2 <- Tcell_subsets %>% select(Module, Class, Sample, Delta.Zscore) %>% 
  pivot_wider(names_from = Module, values_from = Delta.Zscore)

pca_calc2 <- mydata_forpca2 %>% select(where(is.numeric)) %>% prcomp(scale = F)

### plotting PCA
pca_calc2 %>% augment(mydata_forpca2) %>%  ggplot(aes(.fittedPC1, .fittedPC2, color = Class))+ 
  geom_point(size = 2)+
  scale_color_manual(values=c(Normal="#92a0b3",AAV2="#de102a", "Gene Therapy"="#ff7821", "xxxxx.xxxxx035"="#F781BF", "xxxxx.xxxxx045"="#b87dc1", Brain= "#4DAF4A"))+
  labs(x="PC1",y="PC2")+
  My_Theme
```

Make heatmap
```{r}
library(ComplexHeatmap)
library(circlize)

### create numerical data matrix with row & column names
t=12
heatmapData2 <- mydata_forpca2 %>% select(-Class) %>%
  remove_rownames %>% column_to_rownames(var="Sample") %>% t

### create annotations (for help see: https://www.biostars.org/p/317349/)
FilenameAnnotations <- Tcell_subsets %>% select(Class,Sample) %>% distinct()
i <- FilenameAnnotations
ann <- data.frame(i$Class)
colnames(ann) <- c("Class")
ann$Class <- factor(ann$Class, levels = c("Normal","AAV2", "Gene Therapy", "xxxxx.xxxxx035", "xxxxx.xxxxx045", "Brain"))
colours <- list("Class" = c(
  "Normal"=	"#92a0b3",
  "AAV2"=	"#de102a",
  "Gene Therapy"=	"#ff7821",
  "xxxxx.xxxxx035"="#F781BF",
  "xxxxx.xxxxx045"="#b87dc1",
  "Brain" = "#4DAF4A"))
colAnn <- HeatmapAnnotation(df = ann,
                            which = 'col',
                            col = colours,
                            border = T,
                            show_annotation_name=F,
                            annotation_name_gp = gpar(fontsize=8, fontface = "bold")
                            )

### select heatmap colours
library(colorRamp2)
col_fun = colorRamp2(c(0, 5, 10), c("#2166AC", "#FDF0D5", "#950018"))

### plot heatmap 
hm2 <- Heatmap(heatmapData2,
              col = col_fun,
              border_gp = gpar(col = "black", lty = 1),
              rect_gp = gpar(col = "black", lwd = 1),
              #column_title = "...",
              #column_title_gp = gpar(fontsize = 10),
              row_title = "T cell subset modules",
              row_title_gp = gpar(fontsize=12, fontface = "bold"),
              show_row_names = T,
              row_names_gp = gpar(fontsize=10, fontface = "bold"),
              show_column_names = F,
              column_title = "Samples",
              column_title_gp = gpar(fontsize=12, fontface = "bold"),
              #column_names_gp = gpar(fontsize=t),
              #column_names_rot = 0,
              #column_names_centered = T,
              #column_names_gp = gpar(fontsize = 10),
              #name = "z score",
              cluster_rows = T,
              cluster_columns = T,
              clustering_method_columns = "ward.D",
              column_split = 2,
              #row_split = 2,
              top_annotation = colAnn,
              heatmap_height = unit(7, "cm"),
              heatmap_width = unit(14, "cm"),
              #right_annotation = row.ha,
              #show_annotation_name = F,
              heatmap_legend_param = list(
                direction = "horizontal",
                #grid_height = 0.5,
                at = c(0, 2, 4, 6, 10),
                labels = c(-0, 2, 4, 6, 10),
                title = "Expression delta Z-score",
                legend_width = unit(2, "cm"),
                title_position = "lefttop")
              )

draw(hm2, heatmap_legend_side = "bottom", annotation_legend_side = "right")
```

### T cell functions

Make scatterplot
```{r}
ggplot(Tcell_functions)+
  geom_point(aes(Module, Delta.Zscore, colour = Class), size = 3, alpha = 0.8)+
  scale_colour_manual(values = c("#92a0b3", "#de102a", "#ff7821", "#F781BF", "#b87dc1", "#4DAF4A"))+
  geom_hline(aes(yintercept = 0), linewidth = 0.5, linetype = "dashed")+
  scale_y_continuous(limits=c(-2.5, 6), breaks=c(-2.5, 0, 2.5, 5))+
  labs(colour = "Group",
       y = "Delta Zscore cf. average of normal liver", 
       x = "T cell function modules") +
  My_Theme + theme(axis.text.x = element_text(angle = -45, hjust=0))
```


Make PCA plot
```{r}
library(broom)

### calculating PCA data
mydata_forpca3 <- Tcell_functions %>% select(Module, Class, Sample, Delta.Zscore) %>% 
  pivot_wider(names_from = Module, values_from = Delta.Zscore)

pca_calc3 <- mydata_forpca3 %>% select(where(is.numeric)) %>% prcomp(scale = F)

### plotting PCA
pca_calc3 %>% augment(mydata_forpca3) %>%  ggplot(aes(.fittedPC1, .fittedPC2, color = Class))+ 
  geom_point(size = 2)+
  scale_color_manual(values=c(Normal="#92a0b3",AAV2="#de102a", "Gene Therapy"="#ff7821", "xxxxx.xxxxx035"="#F781BF", "xxxxx.xxxxx045"="#b87dc1", Brain= "#4DAF4A"))+
  labs(x="PC1",y="PC2")+
  My_Theme
```

Make heatmap
```{r}
library(ComplexHeatmap)
library(circlize)

### create numerical data matrix with row & column names
t=12
heatmapData3 <- mydata_forpca3 %>% select(-Class) %>%
  remove_rownames %>% column_to_rownames(var="Sample") %>% t

### create annotations (for help see: https://www.biostars.org/p/317349/)
FilenameAnnotations <- Tcell_functions %>% select(Class,Sample) %>% distinct()
i <- FilenameAnnotations
ann <- data.frame(i$Class)
colnames(ann) <- c("Class")
ann$Class <- factor(ann$Class, levels = c("Normal","AAV2", "Gene Therapy", "xxxxx.xxxxx035", "xxxxx.xxxxx045", "Brain"))
colours <- list("Class" = c(
  "Normal"=	"#92a0b3",
  "AAV2"=	"#de102a",
  "Gene Therapy"=	"#ff7821",
  "xxxxx.xxxxx035"="#F781BF",
  "xxxxx.xxxxx045"="#b87dc1",
  "Brain" = "#4DAF4A"))
colAnn <- HeatmapAnnotation(df = ann,
                            which = 'col',
                            col = colours,
                            border = T,
                            show_annotation_name=F,
                            annotation_name_gp = gpar(fontsize=8, fontface = "bold")
                            )

### select heatmap colours
library(colorRamp2)
col_fun = colorRamp2(c(0, 5, 10), c("#2166AC", "#FDF0D5", "#950018"))

### plot heatmap 
hm3 <- Heatmap(heatmapData3,
              col = col_fun,
              border_gp = gpar(col = "black", lty = 1),
              rect_gp = gpar(col = "black", lwd = 1),
              #column_title = "...",
              #column_title_gp = gpar(fontsize = 10),
              row_title = "T cell function modules",
              row_title_gp = gpar(fontsize=12, fontface = "bold"),
              show_row_names = T,
              row_names_gp = gpar(fontsize=10, fontface = "bold"),
              show_column_names = F,
              column_title = "Samples",
              column_title_gp = gpar(fontsize=12, fontface = "bold"),
              #column_names_gp = gpar(fontsize=t),
              #column_names_rot = 0,
              #column_names_centered = T,
              #column_names_gp = gpar(fontsize = 10),
              #name = "z score",
              cluster_rows = T,
              cluster_columns = T,
              clustering_method_columns = "ward.D",
              column_split = 2,
              #row_split = 2,
              top_annotation = colAnn,
              heatmap_height = unit(7, "cm"),
              heatmap_width = unit(14, "cm"),
              #right_annotation = row.ha,
              #show_annotation_name = F,
              heatmap_legend_param = list(
                direction = "horizontal",
                #grid_height = 0.5,
                at = c(0, 2, 4, 6, 10),
                labels = c(-0, 2, 4, 6, 10),
                title = "Expression delta Z-score",
                legend_width = unit(2, "cm"),
                title_position = "lefttop")
              )

draw(hm3, heatmap_legend_side = "bottom", annotation_legend_side = "right")
```


### Myeloid cell functions
Make scatterplot
```{r}
ggplot(myeloid_functions)+
  geom_point(aes(Module, Delta.Zscore, colour = Class), size = 3, alpha = 0.8)+
  scale_colour_manual(values = c("#92a0b3", "#de102a", "#ff7821", "#F781BF", "#b87dc1", "#4DAF4A"))+
  geom_hline(aes(yintercept = 0), linewidth = 0.5, linetype = "dashed")+
  scale_y_continuous(limits=c(-2.5, 6), breaks=c(-2.5, 0, 2.5, 5))+
  labs(colour = "Group",
       y = "Delta Zscore cf. average of normal liver", 
       x = "Myeloid cell function modules") +
  My_Theme + theme(axis.text.x = element_text(angle = -45, hjust=0))
```


Make PCA plot
```{r}
library(broom)

### calculating PCA data
mydata_forpca4 <- myeloid_functions %>% select(Module, Class, Sample, Delta.Zscore) %>% 
  pivot_wider(names_from = Module, values_from = Delta.Zscore)

pca_calc4 <- mydata_forpca4 %>% select(where(is.numeric)) %>% prcomp(scale = F)

### plotting PCA
pca_calc4 %>% augment(mydata_forpca4) %>%  ggplot(aes(.fittedPC1, .fittedPC2, color = Class))+ 
  geom_point(size = 2)+
  scale_color_manual(values=c(Normal="#92a0b3",AAV2="#de102a", "Gene Therapy"="#ff7821", "xxxxx.xxxxx035"="#F781BF", "xxxxx.xxxxx045"="#b87dc1", Brain= "#4DAF4A"))+
  labs(x="PC1",y="PC2")+
  My_Theme
```

Make heatmap
```{r}
library(ComplexHeatmap)
library(circlize)

### create numerical data matrix with row & column names
t=12
heatmapData4 <- mydata_forpca4 %>% select(-Class) %>%
  remove_rownames %>% column_to_rownames(var="Sample") %>% t

### create annotations (for help see: https://www.biostars.org/p/317349/)
FilenameAnnotations <- myeloid_functions %>% select(Class,Sample) %>% distinct()
i <- FilenameAnnotations
ann <- data.frame(i$Class)
colnames(ann) <- c("Class")
ann$Class <- factor(ann$Class, levels = c("Normal","AAV2", "Gene Therapy", "xxxxx.xxxxx035", "xxxxx.xxxxx045", "Brain"))
colours <- list("Class" = c(
  "Normal"=	"#92a0b3",
  "AAV2"=	"#de102a",
  "Gene Therapy"=	"#ff7821",
  "xxxxx.xxxxx035"="#F781BF",
  "xxxxx.xxxxx045"="#b87dc1",
  "Brain" = "#4DAF4A"))
colAnn <- HeatmapAnnotation(df = ann,
                            which = 'col',
                            col = colours,
                            border = T,
                            show_annotation_name=F,
                            annotation_name_gp = gpar(fontsize=8, fontface = "bold")
                            )

### select heatmap colours
library(colorRamp2)
col_fun = colorRamp2(c(0, 5, 10), c("#2166AC", "#FDF0D5", "#950018"))

### plot heatmap 
hm4 <- Heatmap(heatmapData4,
              col = col_fun,
              border_gp = gpar(col = "black", lty = 1),
              rect_gp = gpar(col = "black", lwd = 1),
              #column_title = "...",
              #column_title_gp = gpar(fontsize = 10),
              row_title = "Myeloid cell function modules",
              row_title_gp = gpar(fontsize=12, fontface = "bold"),
              show_row_names = T,
              row_names_gp = gpar(fontsize=10, fontface = "bold"),
              show_column_names = F,
              column_title = "Samples",
              column_title_gp = gpar(fontsize=12, fontface = "bold"),
              #column_names_gp = gpar(fontsize=t),
              #column_names_rot = 0,
              #column_names_centered = T,
              #column_names_gp = gpar(fontsize = 10),
              #name = "z score",
              cluster_rows = T,
              cluster_columns = T,
              clustering_method_columns = "ward.D",
              column_split = 2,
              #row_split = 2,
              top_annotation = colAnn,
              heatmap_height = unit(7, "cm"),
              heatmap_width = unit(14, "cm"),
              #right_annotation = row.ha,
              #show_annotation_name = F,
              heatmap_legend_param = list(
                direction = "horizontal",
                #grid_height = 0.5,
                at = c(0, 2, 4, 6, 10),
                labels = c(-0, 2, 4, 6, 10),
                title = "Expression delta Z-score",
                legend_width = unit(2, "cm"),
                title_position = "lefttop")
              )

draw(hm4, heatmap_legend_side = "bottom", annotation_legend_side = "right")
```
